---
title: Getting Started - Data Cleaning
author: Kent Orr
date: '2022-05-21'
slug: getting-started-data-cleaning
categories: []
tags: []
---

Every data project starts with a broad overview of the information at hand and formation of research questions or business objectives. In this case the questions are conveniently provided. Below is a simple overview of the research questions and provided data sets. 

```{r setup}
library(data.table)
library(ggplot2)

for (i in list.files("/home/kent/Documents/oushowcase/content/data")) {
  if (!grepl(".csv", i)) return()
  nm = gsub(".csv", "", i) 
  assign(nm, fread(paste0("/home/kent/Documents/oushowcase/content/data/",i)) |> 
           janitor::clean_names()
         )
}
```

The state has launched the ‘Smart Start’ Program and the university is being asked to review entry level courses. The following hypotheses are to be tested:

* What are our overall success rates (C or better)?  
* What student information helps predict success in their courses?   
* Are success rates improving or getting worse?  
* Are some of their instructors better than others?  
* Does teaching frequently the course improve instructor performance over time?  

## The Data

Common keys:

Before going any further, it is plain that the data provided resembles a relational database. I want to know how the databases interact. I'll plot the common keys between tables and give a brief, single sentence overview of the content. This document will be used as a reference during investigation.

```{r}
#| echo: true
#| code-fold: true

lapply(c("acad_plan", "class_instructors", "class_inventory", "course_enrollments", "student_details"),
       \(x) {
         data.table(table = x, cols = names(get(x)))
       }) |>
  rbindlist() -> tables

tables[, merge(.SD, tables[,.N, cols])][,common := N>1][] |>
  ggplot(aes(x = reorder(table, -N), y = cols, fill = common, label = table)) + 
  geom_tile(color = "black") + 
  theme_bw() +
  labs(title = "Overview of Data Keys and Columns",
       x = "Tables", y = "Columns", fill = "Is A Common Key") + 
  theme(legend.position = "bottom") + 
  scale_fill_viridis_d(option = "D") 

```


#### class_instructors

links instructors with a course number and term.

```{r}
class_instructors
```

#### course_enrollments

link students to specific course numbers per term, along with some metadata such as hours carried. 

```{r}
course_enrollments
```

#### class_inventory

metadata for courses

```{r}
class_inventory
```

#### student_details

information about a student, demographic data independent of university experience

```{r}
student_details
```

#### acad_plan

metadata on various academic plans

```{r}
acad_plan
```