---
title: Are some instructors better than others?
author: Kent Orr
date: '2022-05-23'
slug: are-some-instructors-better-than-others
categories: []
tags: []
---

```{r setup, include=FALSE}
library(data.table)
library(ggplot2)
source("~/Documents/oushowcase/content/scripts/clean.R")
```

## Defining the Problem

Do some instructors have significantly higher success rates among the students in their courses than others? We will merge the `course_enrollments` with `class_instructors`. From there we will determine success ratios per instructor and compare. 

```{r}
is <- merge(course_enrollments[,.(term_code, class_nbr, 
                                  student_id_number, official_grade,
                                  success)],
            class_instructors, on = c("term_code", "class_nbr"))
```



```{r}
iss <- is[,.N, .(instructor_id_number, term_code, class_nbr, success)
   ][,percent := N/sum(N), .(instructor_id_number, term_code, class_nbr)
     ][]
```

```{r}
iss[(success)] |>
  ggplot(aes(x = reorder(as.factor(instructor_id_number), percent), y = percent)) + 
  geom_boxplot() + 
  theme_bw() + 
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  coord_flip() + 
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Percent of Successful Students",
       x = "Instructors",
       title = "Distribution of Successful Students by Instructor")
```

Do we really need a statistical test to determine if some instructors perform better than others?

I will merge the median success rate for each instructor to the performance ratio data and create two groups, the top 50% and bottom 50%, then use a t-test to test for a difference of means. 

```{r}
iss[(success)][,.(md = median(percent)), instructor_id_number
               ][order(md)
                 ][,split := .I%/%(.N/2), .I
                   ][,.(instructor_id_number, split = split > 0)] |>
  merge(iss[(success)], y = _) -> iss_split

iss_split |>
  ggplot(aes(x = percent, fill = split)) +
  geom_density(alpha = .3) + 
  theme_classic() +
  labs(y = "Success Rate", x = "Probability Density", fill = "Top 50%")
```

If I squint real hard, it looks normal enough for a parametric test. 

```{r}
t.test(iss_split[(split)]$percent, iss_split[!(split)]$percent)
```

But to cover bases I'll apply

```{r}
wilcox.test(iss_split[(split)]$percent, iss_split[!(split)]$percent)
```

There is strong evidence that some instructors' courses have significantly higher success rates than others. 